---
title: "Introduction to R Part 4"
output: html_notebook
---

#### R in a nutshell

- Statistical programming environments
- Originally designed and implemented by statisticians
- Widely popular due to its extensive collection of community-contributed packages
- Quickly gaining places among traditional proprietary tools such as SAS and STATA for data analytics

#### Learning Objectives

- Understand basic programming concepts: variables, assignment, functions, loops, conditions
- Understand core R concepts: data loading, data types, data access, libraries
- Understand advanced R concepts: data manipulation, visualization
- Understand HPC concepts: running R codes on the Palmetto supercomputer via batch submission scripts

#### Materials on this notebook is based on two lessons by Software Carpentry and Data Carpentry:

- Introduction to Programming using R
- Data Analysis and Visualization in R for Ecology

# Running R on a Supercomputer (if not running RStudio from inside Palmetto)

- Log into to Palmetto via JupyterHub interface: https://www.palmetto.clemson.edu/jupyter or MobaXterm
- Request an allocation with the following parameter: 1 chunk, 8 cores, 8gb memory, no gpu, 1 hour
- Create a directory called **intro-to-r**
- Create a directory called **data** inside **intro-to-r**
- Upload the .csv files from your machine into **data** on Palmetto using JupyterHub interface

```{r}
library(knitr)
knit_hooks$set(write_chunk = function(before, options, envir) {
    if (before) {
      fileConn<-file(paste0(options$label,".R") )
      writeLines(options$code, fileConn)
      close(fileConn)
    }
})
```

- Create a new text file named **analyze-1.R** inside **intro-to-r**

```{r analyze-1, eval=FALSE, include=FALSE, write_chunk=TRUE}
setwd("/home/lngo/intro-to-r/")
surveys <- read.csv("data/combined.csv")

library(dplyr)

surveys_complete <- surveys %>%
  filter(species_id != "",         # remove missing species_id
         !is.na(weight),           # remove missing weight
         !is.na(hindfoot_length),  # remove missing hindfoot_length
         sex != "")                # remove missing sex
    
## Extract the most common species_id
species_counts <- surveys_complete %>%
  group_by(species_id) %>%
  tally %>%
  filter(n >= 50)
    
## Only keep the most common species
surveys_complete <- surveys_complete %>%
  filter(species_id %in% species_counts$species_id)

yearly_sex_counts <- surveys_complete %>%
  group_by(year, species_id, sex) %>%
  tally
    
library(ggplot2)

my_plot <- ggplot(data = yearly_sex_counts, aes(x = year, y = n, color = sex, group = sex)) +
    geom_line() +
    facet_wrap(~ species_id) +
    labs(title = 'Observed species in time',
        x = 'Year of observation',
        y = 'Number of species') +
    theme_bw() +
    theme(axis.text.x = element_text(colour="grey20", size=12, angle=90, hjust=.5, vjust=.5),
                        axis.text.y = element_text(colour="grey20", size=12),
          text=element_text(size=16, family="Arial"))

current_dir <- getwd()
output_graph_dir <- 'output_graph'

if (!file.exists(output_graph_dir)){
    dir.create(file.path(current_dir, output_graph_dir))
} else {
    print ("Directory already exists")
}

ggsave(file.path(current_dir, output_graph_dir,"yearly_sex_counts_2.png"), 
       my_plot, width=15, height=10)
```

- Upload **analyze-1.R** into **intro-to-r** on Palmetto
- Test run **analyze-1.R**

```{bash}
R CMD BATCH analyze-1.R
```

- Create a new text file named **analyze-2.R** inside **intro-to-r**

```{r analyze-2, eval=FALSE, include=FALSE, write_chunk=TRUE}
args = commandArgs(trailingOnly=TRUE)

print (args)

setwd(args[1])
surveys <- read.csv(args[2])

library(dplyr)

surveys_complete <- surveys %>%
  filter(species_id != "",         # remove missing species_id
         !is.na(weight),           # remove missing weight
         !is.na(hindfoot_length),  # remove missing hindfoot_length
         sex != "")                # remove missing sex
    
## Extract the most common species_id
species_counts <- surveys_complete %>%
  group_by(species_id) %>%
  tally %>%
  filter(n >= 50)
    
## Only keep the most common species
surveys_complete <- surveys_complete %>%
  filter(species_id %in% species_counts$species_id)

yearly_sex_counts <- surveys_complete %>%
  group_by(year, species_id, sex) %>%
  tally
    
library(ggplot2)

my_plot <- ggplot(data = yearly_sex_counts, aes(x = year, y = n, color = sex, group = sex)) +
    geom_line() +
    facet_wrap(~ species_id) +
    labs(title = 'Observed species in time',
        x = 'Year of observation',
        y = 'Number of species') +
    theme_bw() +
    theme(axis.text.x = element_text(colour="grey20", size=12, angle=90, hjust=.5, vjust=.5),
                        axis.text.y = element_text(colour="grey20", size=12),
          text=element_text(size=16, family="Arial"))

current_dir <- getwd()
output_graph_dir <- 'output_graph'

if (!file.exists(output_graph_dir)){
    dir.create(file.path(current_dir, output_graph_dir))
} else {
    print ("Directory already exists")
}

ggsave(file.path(current_dir, output_graph_dir, args[3]), 
       my_plot, width=15, height=10)
```

- Upload **analyze-2.R** into **intro-to-r** on Palmetto
- Test run **analyze-2.R**

```{bash}
Rscript analyze-2.R /home/$USER/workshop-r-intro-data-science/ data/combined.csv years_sex_counts-3.png
```

- Create a new text file named **analyze-3.R**

```{r analyze-3, eval=FALSE, include=FALSE, write_chunk=TRUE}
#!/usr/bin/env /usr/local/share/jupyterhub/env/R/lib/R/bin/Rscript
args = commandArgs(trailingOnly=TRUE)

print (args)

setwd(args[1])
surveys <- read.csv(args[2])

library(dplyr)

surveys_complete <- surveys %>%
  filter(species_id != "",         # remove missing species_id
         !is.na(weight),           # remove missing weight
         !is.na(hindfoot_length),  # remove missing hindfoot_length
         sex != "")                # remove missing sex
    
## Extract the most common species_id
species_counts <- surveys_complete %>%
  group_by(species_id) %>%
  tally %>%
  filter(n >= 50)
    
## Only keep the most common species
surveys_complete <- surveys_complete %>%
  filter(species_id %in% species_counts$species_id)

yearly_sex_counts <- surveys_complete %>%
  group_by(year, species_id, sex) %>%
  tally
    
library(ggplot2)

my_plot <- ggplot(data = yearly_sex_counts, aes(x = year, y = n, color = sex, group = sex)) +
    geom_line() +
    facet_wrap(~ species_id) +
    labs(title = 'Observed species in time',
        x = 'Year of observation',
        y = 'Number of species') +
    theme_bw() +
    theme(axis.text.x = element_text(colour="grey20", size=12, angle=90, hjust=.5, vjust=.5),
                        axis.text.y = element_text(colour="grey20", size=12),
          text=element_text(size=16, family="Arial"))

current_dir <- getwd()
output_graph_dir <- 'output_graph'

if (!file.exists(output_graph_dir)){
    dir.create(file.path(current_dir, output_graph_dir))
} else {
    print ("Directory already exists")
}

ggsave(file.path(current_dir, output_graph_dir, args[3]), 
       my_plot, width=15, height=10)
```

- Upload **analyze-3.R** into **intro-to-r** on Palmetto
- Test run **analyze-3.R**

```{bash}
$ chmod 755 /home/lngo/intro-to-r/analyze-3.R
$ /home/lngo/intro-to-r/analyze-3.R /home/lngo/intro-to-r/ data/combined.csv years_sex_counts-4.png
```

- Create a submission script for Palmetto named **analyze.pbs** in **intro-to-r**:

```{r}
library(knitr)
knit_hooks$set(write_script = function(before, options, envir) {
    if (before) {
      fileConn<-file(paste0(options$label,".pbs") )
      writeLines(options$code, fileConn)
      close(fileConn)
    }
})
```

```{r analyze, eval=FALSE, include=FALSE, write_script=TRUE}
#!/bin/bash

#PBS -N surveys_data
#PBS -l select=1:ncpus=8:mem=8gb
#PBS -l walltime=00:05:00
#PBS -j oe

cd $PBS_O_WORKDIR

rm output_graph/*
./analyze-3.R $HOME/workshop-r-intro-data-science/data/combined.csv years_sex_counts-pbs.png
```

- Upload **analyze.pbs** into **intro-to-r** on Palmetto
- Open a terminal on JupyterHub interface
- Login to the head node using the command *ssh login001*
- Change directory to **intro-to-r**
- Submit the **analyze.pbs** script
- Observe the process

```{bash}
$ cd /home/lngo/intro-to-r
$ qsub analyze.pbs
$ qstat -anu lngo
```