---
title: "Introduction to R Part 1"
output: html_notebook
---
 
#### R in a nutshell

- Statistical programming environments
- Originally designed and implemented by statisticians
- Widely popular due to its extensive collection of community-contributed packages
- Quickly gaining places among traditional proprietary tools such as SAS and STATA for data analytics

#### Learning Objectives

- Understand basic programming concepts: variables, assignment, functions, loops, conditions
- Understand core R concepts: data loading, data types, data access, libraries
- Understand advanced R concepts: data manipulation, visualization

#### Materials on this notebook is based on two lessons by Software Carpentry and Data Carpentry:

- Introduction to Programming using R
- Data Analysis and Visualization in R for Ecology

# Introduction to Programming using R

## Where am I?

```{r}
getwd()
```

## Variables and assignment

```{r}
read.csv("data/combined.csv")
```

- *variable* : label, name, identifier ...
- *value*    : the actual content represented by a *variable*
- *assignment* : the act of assigning a value to a variable
- R's assignment notation:  *variable* <- *value*


```{r}
x <- 2
```

```{r}
x
```

```{r}
weight_kg <- 97
```

```{r}
weight_lb <- weight_kg * 2.2
```

```{r}
weight_lb
```

```{r}
weight_kg <- 90
```

```{r}
weight_lb <- weight_kg * 2.2
weight_lb
```

*Read data into a variable:*


```{r}
surveys <- read.csv("data/combined.csv")
```


```{r}
head(surveys)
```

*Header is TRUE or Header is FALSE, that is the question!*

```{r}
surveys <- read.csv("data/combined.csv", header = FALSE)
head(surveys)
```

```{r}
surveys <- read.csv("data/combined.csv")
head(surveys)
```

*How to get help?*

```{r}
?read.csv
```

## Data Types

- Data frames are the *de facto* data structure for R's tabular data, and conceptionally equivalent to an Excel spreadsheet but is more powerful and versatile.
- Matrices (multi-dimensional) and vectors (one dimension) are also available for computational purposes. 
- Data frames represents a table whose columns are vectors with same length but possible different data types

```{r}
class(surveys)
```

*Structure of a data frame*

```{r}
str(surveys)
```

```{r}
summary(surveys)
```

*Size of a data frame*

```{r}
dim(surveys)
```

```{r}
nrow(surveys)
```

```{r}
ncol(surveys)
```

*Content of a data frame*

```{r}
head(surveys)
```

```{r}
head(surveys, n=10)
```

```{r}
tail(surveys)
```

```{r}
tail(surveys, n=10)
```

*Names*

```{r}
names(surveys)
```

```{r}
surveys_colnames <- names(surveys)
```

```{r}
surveys_colnames
```

```{r}
surveys_rownames <- rownames(surveys)
str(surveys_rownames)
```

## Data frames access: indexing and subsetting

Similar to an Excel spreadsheet, we can extract specific data from a dataframe via 'coordinates': row/column combinations

Accessing a single element

```{r}
surveys[1,1]
```

```{r}
surveys[1,2]
```

Accessing a block of elements

```{r}
surveys[1:5,2]
```

```{r}
surveys[2,3:7]
```

```{r}
surveys[1:5,3:7]
```

Accessing scattered groups of elements

```{r}
?c
```

```{r}
surveys[c(2:4,6:7),]
```

Excluding data with the `-` notation:

```{r}
surveys[1:5, -3]
```

Accessing columns by names:

```{r}
surveys[1:5,"month"]
```

```{r}
surveys[["month"]][1:5]
```


```{r}
surveys$month[1:5]
```

** Challenge: **

- Create a data frame containing on observations from row 200 to the end of the `surveys` data set
- Create a data frame containing the row that is in the middle of the data frame. Store the content in a variable named `surveys_middle`.
- Combine `nrow` with the `-` notation to reproduce the behavior of `head(surveys)`

```{r}

```

```{r}

```

```{r}

```

## Factors

- Special class, representing categorical data
- Can be ordered or unordered
- Stored as integers with labels (text) associated with these unique integers
- Looked and behave like character vectors but are integers under the hood
- Once created, a `factor` object can only contain a pre-defined set of values, known as *levels*. 
- *Levels* are sorted alphabetically by default. 

```{r}
str(surveys)
```

```{r}
levels(surveys$sex)
```

```{r}
nlevels(surveys$sex)
```

Converting factors:

```{r}
as.character(surveys$sex)
```

```{r}
f <- factor(c(1990,1983,1977,1998,1990))
```

```{r}
f
```

```{r}
as.numeric(f) #incorrect
```

```{r}
as.numeric(as.character(f)) #works
```

```{r}
as.numeric(levels(f))[f] #recommended
```

Renaming factors:

```{r}
plot(surveys$sex)
```

```{r}
sex <- surveys$sex
```

```{r}
levels(sex)
```

```{r}
levels(sex)[1] <- "missing"
```

```{r}
plot(sex)
```

Using `stringsAsFactors=FALSE`

```{r}
surveys <- read.csv('data/combined.csv', stringsAsFactors = TRUE)
str(surveys)
```

```{r}
surveys <- read.csv('data/combined.csv', stringsAsFactors = FALSE)
str(surveys)
```