---
title: "Introduction to R Part 2"
output: html_notebook
---

#### R in a nutshell

- Statistical programming environments
- Originally designed and implemented by statisticians
- Widely popular due to its extensive collection of community-contributed packages
- Quickly gaining places among traditional proprietary tools such as SAS and STATA for data analytics

#### Learning Objectives

- Understand basic programming concepts: variables, assignment, functions, loops, conditions
- Understand core R concepts: data loading, data types, data access, libraries
- Understand advanced R concepts: data manipulation, visualization

#### Materials on this notebook is based on two lessons by Software Carpentry and Data Carpentry:

- Introduction to Programming using R
- Data Analysis and Visualization in R for Ecology

# Introduction to Data Manipulation using R

## Where am I?


```{r}
getwd()
```

```{r}
surveys <- read.csv("data/combined.csv")
head(surveys)
```

## Data Frames Manipulation

```{r}
if (!require('dplyr', character.only = TRUE)){
  install.packages('dplyr', dep = TRUE)
  if (!require('dplyr', character.only = TRUE)){
    print ('Package not found')
  }
}
```

** Common `dplyr` functions: **
- `select()`
- `filter()`
- `mutate()`
- `groupby()`
- `summarize()`
- `%>%`

** Cheatsheet: **
http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

** Selecting columns and filtering rows **

```{r}
select(surveys, plot_id, species_id, weight)
```


```{r}
filter(surveys, year == 1995)
```

** Pipes: combining multile select and filter actions **

```{r}
surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)
```

```{r}
surveys_sml <- surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)

surveys_sml
```

** Challenge: **

Using pipes, subset the `survey` data to include individuals collected before 1995 and retain only the columns `year`, `sex`, and `weight`. 

```{r}

```

** Mutate: create new columns based on existing columns **


```{r}
surveys %>%
  mutate(weight_kg = weight / 1000) %>%
  head  
```

```{r}
surveys %>%
  filter(!is.na(weight)) %>%
  mutate(weight_kg = weight / 1000) %>%
  head
```

- `is.na`: determines whether something is NA (not available - missing values)
- `!`: negates a logical value

** Challenge: **

Create a new data frame from the survey data that meets the following criteria: 

contains only the species_id column and a new column called hindfoot_half containing values that are half the hindfoot_length values. In this hindfoot_half column, there are no NAs and all values are less than 30.

```{r}

```


** Split-apply-combine data analysis and the summarize() function **

- split data into groups
- apply some analysis to each group
- combine the results

** `group_by()` and `summarize()`: **

```{r}
surveys %>%
  group_by(sex) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))
```

```{r}
surveys %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))
```

- `NaN`: not a number
- Need filtering to remove missing values

```{r}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight))
```

```{r}
x <- surveys %>%
       filter(!is.na(weight)) %>%
       group_by(sex, species_id) %>%
       summarize(mean_weight = mean(weight))
str(x)
```

- If you want to display more data, you use the `print()` function at the end of your chain with the argument `n` specifying the number of rows to display:

```{r}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight)) %>%
  print(n = 15)
```

Summarization on multiple variables at the same time is also possible

```{r}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight))
```

** Tallying: simply counting things: **

```{r}
surveys %>%
  group_by(sex) %>%
  tally
```

** Challenge: **

- How many individuals were caught in each plot_type surveyed?
- Use `group_by()` and `summarize()` to find the mean, min, and max hindfoot length for each species (using `species_id`).

```{r}

```

```{r}

```


** Exporting data to file: **

- Write cleaned data to file, so that data cleaning process does not have to redone
- Output data should be stored in different location from original raw data

Conditional statement:

```
if (condition is true){
    do something
} else {
    do something else
}
```

```{r}
current_dir <- getwd()
output_data_dir <- 'output_data'

if (!file.exists(output_data_dir)){
    dir.create(file.path(current_dir, output_data_dir))
} else {
    print ("Directory already exists")
}
```

```{r}
?dir.create
```


```{r}
?file.path
```

Create a clean data set with missing observations removed

```{r}
surveys_complete <- surveys %>%
  filter(species_id != "",         # remove missing species_id
         !is.na(weight),           # remove missing weight
         !is.na(hindfoot_length),  # remove missing hindfoot_length
         sex != "")                # remove missing sex
```

Additional filters: Remove rare species (less than 50 observations)

- Create index of rare species
- Filter rare species from the cleaned data set

```{r}
## Extract the most common species_id
species_counts <- surveys_complete %>%
  group_by(species_id) %>%
  tally %>%
  filter(n >= 50)
```

```{r}
## Only keep the most common species
surveys_complete <- surveys_complete %>%
  filter(species_id %in% species_counts$species_id)
```

Write data to file

```{r}
write.csv(surveys_complete, file = file.path(output_data_dir, "surveys_complete.csv"),
          row.names=FALSE)
```